% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}



\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Practical Data Stewardship for Salmon Biologists},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Practical Data Stewardship for Salmon Biologists}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{DRAFT A Blueprint for Domain-Specific Best Practices}
\author{Brett Johnson \and Scott Akenhead \and Katie
Barnas \and Jennifer Bayer \and Tomas Bird \and Bruno
Carturan \and Samuel Cimino \and Graeme Diack \and Lara
Erikson \and Eric Hertz \and Nancy Leonard \and Catherine
Michielsens \and Fiona Martens \and Emily Lescak \and Stephanie
Peacock \and Gottfried Pestal \and Matt Jones \and Mark Saunders \and Yi
Xu}
\date{}
\begin{document}
\maketitle


\textbf{Keywords:} Salmon data stewardship, data interoperability, FAIR
principles, persistent identifiers (PIDs), controlled vocabularies,
metadata standards, application programming Interface (API), data
citation, ontology development

\subsubsection{Abstract}\label{abstract}

Fisheries management, research, and broader conservation efforts
increasingly generate vast, diverse datasets needed for timely
decision-making. Yet these data remain fragmented across jurisdictions,
disciplines, and legacy infrastructure, limiting responsive management.
The lack of access, interoperability, and reuse of data are major
impediments to better science; data stewardship---the set of social and
technical practices that ensure data are responsibly created, managed,
shared, and reused over time---is the solution. Biologists are
increasingly taking on data stewardship responsibilities to address
these challenges, often without clear guidance, training, or support.
Shared, community-agreed practices for implementing domain-specific data
standards are needed to move beyond generic data management guidance
toward fit-for-purpose tools and workflows. To address this gap---and to
show how other communities can do so---we develop seven practices for
salmon data stewardship. Through a case study of a historically
important sockeye salmon productivity analysis that spanned Pacific
Coast jurisdictions, we highlight how clearly defined data stewardship
practices could improve data integration, management efficacy, and
reduce the overall cost of updating the analysis annually. We provide
practical guidance for those transitioning into these essential
stewardship roles, including salmon domain specific data tools,
templates, and examples. These practices integrate both sociocultural
and technical approaches to ensure data meet modern open science
principles and respect Indigenous Data Sovereignty. With a foundation of
shared practices, data stewards will enable faster, more transparent
decision-making. They will expand the reuse of cross-jurisdictional
datasets and support machine-actionable data with high-quality metadata
and consistent semantics. Together, these outcomes strengthen the
management and conservation of salmon populations and the ecosystems
they inhabit, with broader relevance to other data-rich fisheries
domains.

\subsection{The Data Stewardship
Challenge}\label{the-data-stewardship-challenge}

Effective fisheries management relies on robust, accessible data on
abundance, health, and environmental conditions integrated over large
spatial areas and fine temporal scales. Yet across many fisheries, data
remain fragmented, inconsistently measured, and incomplete across time,
space, and life history stages, which limits their value for research,
hypothesis testing, and decision-making Marmorek et al. (2011); Inman et
al. (2021); Diack et al. (2024). Collections are often managed by
diverse agencies and organizations in isolation, undermining timely,
integrated analyses necessary for adaptive management. The absence of
organizational support and dedicated roles for data stewardship further
compounds this problem, with critical tasks often treated as ad hoc
responsibilities rather than core management functions. These issues
limit the utility of these data for research and management.

\subsubsection{Fragmented Landscapes}\label{fragmented-landscapes}

For salmon in particular, the integration and mobilization of salmon
data mirrors the complexity of their life cycle, life histories, and
migrations across ecological regions (Groot and Margolis 1991) and
jurisdictional boundaries. Salmon biologists routinely collect
information managed by diverse agencies and organizations, often
isolated within their operational context and without a focus on
interoperability. While localized successes in data coordination
exist---particularly within regional fisheries management offices and
treaty commissions ({``Fraser {River} {Panel} {Data} {Application}''}
2025)---salmon data integrated across agencies for each phase of the
salmon life cycle is uncommon and costly for most programs. Even within
organizations, data can be siloed by data type with freshwater, estuary,
open-ocean, and commercial fishery data each housed in separate data
systems with limited ability to easily re-connect the data through
shared identifiers. The result is a highly fragmented data landscape
that spans agencies, organizations, and monitoring programs, making
integration especially difficult. This lack of coordinated stewardship
has become a significant impediment for adaptive salmon management and
conservation, preventing the full use of existing data to inform timely
and effective decisions.

This fragmentation is compounded by the number of scientific disciplines
involved in piecing together a comprehensive understanding of individual
salmon stocks. Geneticists, oceanographers, freshwater ecologists, stock
assessment biologists, and fisheries managers are just a sample of
disciplines that contribute data to piece together a complete life
history for a salmon population. Each uses their own domain-specific
conventions and workflows, standards, and metadata. This tangle of
disciplinary and organizational fragmentation slows integration, hinders
reproducibility, and delays analyses that could otherwise inform
time-sensitive management decisions, conservation actions, or
restoration plans. Large volumes of data collected through long-term
monitoring programs hold tremendous value, especially for secondary
users---but are often inaccessible due to a lack of time, resources, and
incentives for data producers to publish them (LINDENMAYER et al. 2012).
When critical datasets are hard to find, access, or interpret,
biologists and analysts lose valuable time trying to reconstruct or
harmonize them. This reduces transparency, increases the risk of errors,
delays urgent conservation or management responses, which in turn
undermines the public trust in both the science and the resulting
management decisions. Modernizing these systems will require coordinated
communities of practices, shared practices, and shifting collective
expectations and culture towards shared international data standards and
stewardship practices that accommodate the full disciplinary and
geographic diversity of salmon science.

\subsubsection{Coordinated Data
Stewardship}\label{coordinated-data-stewardship}

Data stewardship offers to face the growing challenges in salmon
management from escalating environmental uncertainties due to climate
change which demands rapid, regionally integrated, and robust data Ward
et al. (2025). The mismatch between fragmented data systems and fixed
administrative and jurisdictional boundaries creates an urgent need for
interoperable, dynamic, multi-scale data stewardship that can adapt to
shifting ecological and management priorities. Despite the scale and
importance of these datasets, biologists who collect and manage salmon
data are ofteno expected to act as de facto data stewards without
training, guidance, incentives, organizational support, or access to
community-agreed best practices. Tasks such as documenting methods,
aligning terminology, formatting for data sharing, and publishing data
are typically performed off the side of a biologist's desk. Done well,
data stewardship is skilled, labor-intensive work---combining domain
expertise, information science, and relationship-building---and it
requires dedicated time and institutional support rather than ad hoc
effort. A lack of organizational support (Diack et al. 2024), training
(Volk, Lucero, and Barnas 2014), and dedicated roles for data management
further relegate critical data stewardship tasks to an \emph{ad hoc}
status. Deferred documentation, governance, and standardization tasks
accumulate as data technical debt---work left undone that compounds over
time, making integration and reanalysis progressively slower and more
error-prone. The absence of clear roles, standards, and
community-endorsed practices leaves even motivated scientists unsure how
to structure their data for future use. As a result, data stewardship is
inconsistent and reactive, and data integration remains a major
bottleneck to adaptive management and ecosystem-scale learning.

Due to a lack of practical resources, both salmon biologists struggle to
align their data with community-agreed principles such as FAIR
(Findable, Accessible, Interoperable, and Reusable) (Wilkinson et al.
2016) and Indigenous Data Sovereignty frameworks like the CARE
principles (Collective Benefit, Authority to Control, Responsibility,
and Ethics) (Carroll, Rodriguez-Lonebear, and Martinez 2019; Jennings et
al. 2023). Adhering to CARE data management principles is all the more
important when it comes to salmon related data given the sociocultural
importance of salmon to First Nations, Tribes, and Indigenous
communities throughout the North Pacific and North Atlantic regions
(Ween and Colombi 2013; Earth Economics 2021). While both FAIR and CARE
present laudable goals with respect to data stewardship, these sets of
principles can sometimes be at odds with each other and implementing
both presents an inherent challenge. Nonetheless, several of the best
practices we propose can help clarify shared objectives and increase
transparency and trust that is essential for integrating both FAIR and
CARE principles. Without clear support and guidance, well-intentioned
practitioners are left with ad hoc approaches that limit reuse and
interoperability. This gap can only be bridged by equipping both data
producers and stewards with tools, support, and organizational backing
to publish interoperable, machine-readable metadata and datasets in
alignment with shared principles.

\subsubsection{Framework for Action}\label{framework-for-action}

In this paper, we provide a framework for translating data stewardship
principles into actionable practices, examples, and workflows to help
salmon biologists improve the usability, traceability, and long-term
impact of their data. We offer seven practices for salmon data
stewardship Table~\ref{tbl-bestpractices} and use a retrospective,
real-world case study of cross-jurisdictional sockeye productivity
synthesis (Peterman and Dorner 2012)to illustrate how the absence of
shared practices made integration time-consuming and keeps annual
updates costly and difficult to this day. To support adoption, we
provide real-world examples from multiple organizations (Appendix 1), a
training roadmap for biologists taking on stewardship responsibilities
(Appendix 2), and a getting-started checklist (Appendix 3). We also map
the seven practices to widely used data-lifecycle model
Table~\ref{tbl-lifecycle} to make adaptation straightforward outside
salmon contexts. Although salmon provide the worked example, the
approach is intended as a transferable blueprint: other scientific
communities can substitute their own vocabularies, repositories,
governance structures, and incentives to turn data stewardship
principles in to practical applications.

A coordinated approach to stewarding salmon data should follow
established open science standards and principles (Johnson and Stap
2024; Wilkinson et al. 2016; Carroll et al. 2020), tailored specifically
within the context of salmon research and management. Our practices
build upon and extend existing data and metadata standards and
infrastructure including Darwin Core, OBIS, schema.org, and OBO Foundry
ontologies, ensuring compatibility with broader biodiversity informatics
infrastructure while avoiding reinventing foundational frameworks.
Achieving meaningful interoperability of data among programs,
organizations and scientific domains demands both breadth and depth.
\textbf{Broad interoperability} integrates diverse scientific domains,
systems, and formats, requiring structured, machine-readable data and
metadata published openly for maximum discoverability. \textbf{Deep
interoperability} demands precise definitions of domain-specific terms
and methods, ensuring data remains meaningful and usable across
contexts. Salmon data stewards can improve outcomes for salmon by
coordinating across boundaries to develop a shared foundation of data
stewardship practices. To address these foundational challenges, we must
establish clear data stewardship roles and practices that span the
entire data lifecycle and salmon lifecycle---from collection and
documentation through integration, long-term preservation, and reuse and
from gravel to estuary and the high seas back to the gravel.

\subsection{Defining Data Stewardship in Salmon
Science}\label{defining-data-stewardship-in-salmon-science}

Data stewardship encompasses the coordinated practices, roles, and
responsibilities necessary to effectively manage, share, and reuse data
throughout its lifecycle (NOAA 2007; Plotkin 2014; Peng et al. 2018). It
includes ensuring data quality, compliance with agreed-upon standards,
and the establishment of clear governance to guide data collection,
documentation, integration, and preservation. However, given the
complexity of the salmon data landscape, stewardship goes beyond mere
technical data management to include actively facilitating
collaboration, communication, and consensus-building among data
producers and users across multiple organizations and jurisdictions.

Data stewardship represents a critical sub-discipline within the broader
field of data science. While data science is often narrowly associated
with machine learning and statistical modeling, we adopt a more
comprehensive view that encompasses how we treat, handle, and represent
data, along with the social and technical information systems that
enable data-intensive science. Data stewardship focuses on the practical
implementation of these principles---ensuring that data infrastructure,
standards, and practices actually serve scientific and management needs
rather than remaining theoretical constructs.

Effective salmon data stewards serve as boundary spanners bridging
agencies, Nations, and disciplines serving as community coordinators,
convening diverse stakeholders to build sustained communities of
practice. This boundary-spanning role is particularly critical in
transboundary contexts where data integration requires navigating
complex jurisdictional and cultural boundaries (Ward et al. 2025). These
responsibilities demand specialized expertise and sustained effort;
without dedicated capacity, standards and workflows tend to degrade, and
technical debt accumulates. By facilitating communication, translating
between different organizational cultures and technical systems, and
maintaining long-term relationships, data stewards create the social
infrastructure necessary for effective cross-boundary data
collaboration.

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, toptitle=1mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, opacityback=0, toprule=.15mm, breakable, rightrule=.15mm, coltitle=black, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Box 1: Critical Functions of Salmon Data Stewards}, bottomrule=.15mm, leftrule=.75mm, bottomtitle=1mm, opacitybacktitle=0.6, titlerule=0mm, left=2mm]

Effective salmon data stewards perform several critical functions:

\begin{itemize}
\item
  \textbf{Technical oversight}: Ensuring metadata completeness,
  adherence to standardized terminologies and vocabularies, and robust
  quality assurance protocols.
\item
  \textbf{Social and organizational facilitation}: Leading stakeholder
  engagement, capacity-building activities, and negotiation of data
  access and sharing agreements, including addressing First Nations,
  Tribes, and Indigenous Peoples' rights and interests in data
  governance.
\item
  \textbf{Organizational advocacy}: Championing the organizational
  recognition of data stewardship roles, promoting sustained investment
  and dedicated resources for data management infrastructure and
  practices.
\item
  \textbf{Implementation and adoption facilitation}: Actively promoting
  data use and ensuring that standards and practices remain practical
  and relevant by maintaining close contact with real-world
  applications. This includes monitoring data utilization, gathering
  feedback from users, and iteratively refining standards based on
  actual implementation challenges to prevent theoretical approaches
  that fail in practice.
\end{itemize}

\end{tcolorbox}

Data stewards can implement FAIR and CARE principles through concrete
technical and governance mechanisms they control, such as documenting
consent constraints and access levels in metadata, using controlled
vocabularies to ensure consistent terminology, and establishing
repository roles that enforce data sovereignty requirements. FAIR is
sometimes interpreted as ``open by default,'' but it does not require
unrestricted openness; rather, it requires that data and metadata be
discoverable and accessible under explicit, machine-readable conditions.
CARE's Authority to Control can therefore place legitimate constraints
on access and reuse, particularly when data are held or governed by
First Nations, Tribes, and Indigenous Peoples. In these cases,
stewardship often means publishing rich, standards-aligned metadata
while implementing governed access, consent-aware reuse conditions, and
culturally appropriate protocols for the data themselves. For example,
stewards can document consent constraints in metadata fields and enforce
access restrictions via repository user roles, ensuring that Indigenous
data sovereignty is respected while maintaining data discoverability and
appropriate reuse (Montenegro 2019; Local Contexts 2025). This
governance approach is particularly critical for sensitive data such as
Traditional Knowledge and sensitive habitat locations, where stewardship
practices must balance open science principles with appropriate access
controls and cultural protocols.

Data stewards play a critical role bridging the gap between biologists
and Information Technology (IT) staff by translating data needs into
application or data system features. A user-centred design approach to
salmon data stewardship is critical and focuses on creating tools that
align with biologists' needs. When data management is separated from
biologists, accountability weakens, and quality issues go unnoticed.
While IT expertise is essential for infrastructure and security,
effective data system design requires IT to act as an enabler, rather
than gatekeeper, provisioning self-serve data infrastructure. The Data
Steward, serving as a translator between IT and biologists, enables
biologists to engage independently with data systems, fostering
ownership and accountability and ultimately improving data quality for
research and management.

Dedicated stewardship roles empower salmon biologists to bridge
disciplinary divides and jurisdictional barriers, transforming
fragmented datasets into cohesive, interoperable resources. By
proactively defining, implementing, and maintaining data standards and
workflows, salmon data stewards create conditions for timely, accurate,
and reproducible analyses. Such stewardship positions salmon biologists
to better inform adaptive management decisions, ultimately strengthening
salmon conservation and resilience.

\subsection{Updating Sockeye Productivity Synthesis: A Case Study for
What Agencies Could Do
Now}\label{updating-sockeye-productivity-synthesis-a-case-study-for-what-agencies-could-do-now}

This case study revisits a comprehensive sockeye productivity dataset
assembled from several agency sources by academic researchers (Peterman
and Dorner 2012). We reflect not on the significant work the research
team accomplished, but rather on the preventable organizational and
technical barriers that impeded their work---and continue to burden data
compilation updates and reuse efforts today. Their study examined
productivity trends across 64 sockeye salmon stocks spanning Washington,
British Columbia (B.C.), and Alaska and found common trends in declining
productivity across the North American range of sockeye salmon (Peterman
and Dorner 2012). This work was important for sockeye conservation and
management in that it pointed to common, broad-scale marine trends in
sockeye salmon productivity, rather than idiosyncratic stock-specific,
freshwater effects---an insight not accessible without a monumental data
integration effort. Attempting to replicate or build upon this analysis
today is an arduous, time-consuming, and error-prone endeavour due to
fragmented data sources, inconsistent formats, and lack of standardized
practices among the key organizations involved: the Washington
Department of Fish and Wildlife (WDFW), Fisheries and Oceans Canada
(DFO), and the Alaska Department of Fish and Game (ADF\&G).

Each section below highlights a key challenge faced by the team and
proposes practical steps based on our best practices
(Table~\ref{tbl-bestpractices}) that data-holding agencies could do to
enable easier integration, validation, and updating of salmon datasets
across jurisdictions and decades. This case study illustrates how
implementing the foundational concepts and practical recommendations
outlined in this paper can transform data stewardship practices within
these organizations. By doing so, they can significantly enhance data
accessibility, quality, and interoperability---ultimately enabling more
efficient and accurate analyses that support salmon conservation and
management.

\subsubsection{Challenge 1: Interpreting the Data --- What do these
numbers actually
mean?}\label{challenge-1-interpreting-the-data-what-do-these-numbers-actually-mean}

Peterman's team frequently worked with datasets that lacked basic
contextual information. Fields such as ``year,'' ``return,'' or ``age
class'' were often undefined or inconsistently used. For example, some
datasets recorded returns by calendar year while others used brood year,
and few included metadata to clarify the distinction. In many cases, the
team had to reconstruct metadata by back-checking against reports or
simulating assumptions (e.g., about age structure) to interpret the data
correctly.

\textbf{Remedies:}

\begin{itemize}
\item
  \textbf{Best Practice 3: Make Data, People, Projects, and Outputs
  Discoverable, Linked and Citable with Persistent Identifiers (PIDs).}
  Assigning PIDs such as digital object identifiers (DOIs) to protocols,
  methods, and people (via ORCIDs) and linking them together using data
  stores and catalogues links data to its provenance and ensures that
  methods, context, and interpretation decisions are traceable.
\item
  \textbf{Best Practice 4. Use Shared Data Models, Vocabularies and
  Metadata to Enable Integration.} To prevent this kind of ambiguity,
  agencies can now adopt internationally recognized metadata schemas
  such as ISO 19115 or Ecological Metadata Language, data models (Darwin
  Core Data Package) to model age and age type data concepts, and use
  controlled vocabularies to restrict the permissible values in the age
  field to calendar year, brood year, or otherwise.
\end{itemize}

\subsubsection{Challenge 2: Accessing and Using the Data --- Where is it
stored, and how do I get
it?}\label{challenge-2-accessing-and-using-the-data-where-is-it-stored-and-how-do-i-get-it}

The Peterman dataset was compiled from multiple files scattered across
email inboxes, regional offices, and grey literature. Data were stored
in inconsistent formats, lacked clear versioning, and were difficult to
discover outside of specific research networks. Even today, no
application programming interface (API) or structured access mechanism
exists to update or query the data programmatically. As a result,
researchers hoping to build on the dataset may have to start from
scratch.

\textbf{Remedies:}

\begin{itemize}
\item
  \textbf{Best Practice 2: Reuse Proven Infrastructure to Save Time and
  Increase Interoperability}\\
  Rather than developing bespoke data catalogues or repositories,
  agencies should adopt existing catalogues used beyond their own
  organization such as the Ocean Biodiversity Information System,
  Zenodo, or the Knowledge Network for Biocomplexity). These are proven
  platforms with a broad user base that support persistent storage,
  discoverability, and interoperability.
\item
  \textbf{Best Practice 5: Store and Analyze Data in Ways That Others
  Can Easily Access, Use, and Trust}\\
  Agencies can use open-access data repositories or their own
  organizational data repositories or catalogues that make data
  discoverable using PIDs and provide programmatic access to data
  possible using APIs.
\end{itemize}

\subsubsection{Challenge 3: Sustaining the Dataset --- Who is
responsible, and why should I
contribute?}\label{challenge-3-sustaining-the-dataset-who-is-responsible-and-why-should-i-contribute}

Once Peterman and his team completed their analysis, no formal plan
existed for sustaining or updating the dataset. Responsibility for
ongoing maintenance fell informally to former students and
collaborators. Despite its national and international relevance, the
dataset was never adopted by an agency as a living product. Moreover,
the original data contributors often lacked incentives, support, or
recognition for their efforts---conditions that persist in many data
environments today.

\textbf{Remedies:}

\begin{itemize}
\item
  \textbf{Best Practice 1: Make Data Governance Explicit to Support
  Trust and Reuse} Agencies should define roles, responsibilities, and
  decision-making processes through formal governance mechanisms such as
  data product charters. Use a Data Management Plan with a
  responsibility matrix such as ``responsible, approver, consulted,
  informed'' (RACI) to clarify governance, assign maintenance
  responsibility, and ensure continuity across staff turnover and
  organizational change.
\item
  \textbf{Best Practice 6: Incentivize and Track Data Sharing and Reuse}
  Visibility, credit, and metrics are critical for motivating data
  sharing. Agencies can embed citation guidance in metadata and track
  dataset reuse through COUNTER-compliant dashboards or DataCite APIs.
\item
  \textbf{Best Practice 7: Build Community Through Co-Development and
  Mutual Benefit} Effective data stewardship requires collaboration
  between biologists, First Nations, Tribes, Indigenous communities,
  managers, and data professionals. Participatory design ensures that
  systems and standards meet user needs and are adopted over time.
  \emph{Practical application:} Facilitate cross-jurisdictional working
  groups to co-develop data standards and align on shared outcomes for
  priority datasets.
\end{itemize}

While the barriers encountered in compiling, interpreting, and
maintaining the data~made the work considerably more difficult than
necessary and hinder efforts by others seeking to extend or build upon
this work, they remain instructive. These challenges are not unique to
Peterman's team---they reflect systemic gaps in data governance,
documentation, infrastructure, and incentives. By adopting the seven
best practices detailed in Table~\ref{tbl-bestpractices}, organizations
and researchers can transform legacy datasets into living resources,
enabling reuse, easing collaboration, and accelerating insight across
the salmon research and management community.

The challenges and solutions demonstrated in this salmon case study
generalize across fisheries and environmental monitoring domains.
Cross-jurisdictional data harmonization, quality assurance and control
patterns, standardized metadata requirements, and long-term archiving
strategies are universal needs that extend far beyond salmon science.
Similar barriers and solutions apply to trawl survey data integration,
invertebrate monitoring programs, and water quality datasets that span
multiple agencies and jurisdictions.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.7035}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.2965}}@{}}
\caption{Best practices and practical applications of salmon data
stewardship}\label{tbl-bestpractices}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Best Practice
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Start here
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Best Practice
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Start here
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1. Make Data Governance Foundational and Explicit to Establish
and Sustain Trust and Reuse.} Establishing clear governance structures
promotes quality, accountability, and alignment with FAIR and CARE
principles enabling trust and long-term stewardship across
multi-organization projects. & - Document roles and responsibilities
using a Data Product Governance Charter and structured responsibility
frameworks.

- Integrate CARE principles to respect First Nations, Tribes, and
Indigenous data rights.

- Form a governance or oversight committee to review data standards,
timelines, and agreements. \\
\textbf{2. Reuse Proven Infrastructure to Save Time and Increase
Interoperability.} Leveraging existing platforms and technologies by
building on and extending them rather than building bespoke solutions
reduces costs and improves long-term interoperability and
sustainability. & - Use domain-specific repositories

- Publish and archive data with KNB or Zenodo. \\
\textbf{3. Make Data, People, Projects, and Outputs Discoverable, Linked
and Citable with Persistent Identifiers.} Persistent identifiers (PIDs)
connect data with researchers, organizations, and outputs---supporting
data citation, reuse, and automated attribution. & - Encourage use of
ORCIDs for researchers.

- Use ROR IDs for organizations.

- Assign DOIs via DataCite for data packages.

- Embed DOIs in dashboards and metadata. \\
\textbf{4. Use Shared Data Models, Vocabularies and Metadata to Enable
Integration.} Common vocabularies, metadata standards, and ontologies
support integration across systems and preserve semantic meaning. & -
Adopt ISO 19115, EML, or DataCite metadata standards.

- Re-use terms defined in Salmon Domain Ontology.

- Model datasets using the Darwin Core Data Package Model. \\
\textbf{5. Store and Analyze Data in Ways That Others Can Easily Access,
Use and Trust.} Structured and accessible data formats ease reusability,
and support integration with analytical tools and applications while
data analyzed or wrangled using programmatic scripts (R, Python etc.)
enable reproducibility and increase trust. & - Provide APIs using
FastAPI, Flask, or Django REST.

- Archive in trusted repositories (e.g., GBIF, FRDR, USGS).

- Write scripts in a programming language to wrangle, transform, and
analyze data.

- Use GitHub to host code for collaboration and transparency and the
GitHub / Zenodo integration for DOI assignment and preservation. \\
\textbf{6. Incentivize and Track Data Sharing and Reuse.} Recognizing
data contributors and tracking reuse promotes a culture of sharing and
supports professional recognition. & - License data with CC-BY 4.0.

- Include citation text and visible credit fields.

- Use COUNTER metrics and DataCite APIs to monitor reuse.

- Encourage dataset citation in references. \\
\textbf{7. Build Community Through Co-Development and Mutual Benefit.}
Engaging users early ensures tools and standards meet real-world needs
and enhances long-term stewardship. & - Participate in RDA Salmon
Interest Group.

- Facilitate workshops for metadata and vocabulary alignment.

- Support community-engaged research with tangible benefits. \\
\end{longtable}

\subsubsection{Metadata governance as a cross-cutting
foundation}\label{metadata-governance-as-a-cross-cutting-foundation}

Unlike the sequential stages of the data lifecycle, metadata governance
operates as a continuous, cross-cutting practice that spans all phases
simultaneously. While data moves through Plan → Obtain → Process →
Preserve → Access → Disposition, metadata governance must be active
throughout, ensuring that documentation, quality control, and
discoverability are maintained at every stage. This cross-cutting nature
means that metadata governance failures at any point can compromise the
entire data stewardship effort, making it a critical foundation rather
than a discrete step in the process.

The lifecycle mapping in Table~\ref{tbl-lifecycle} reveals that data
governance elements appear in every stage: planning metadata
requirements (Plan), documenting collection methods (Obtain),
structuring and validating metadata (Process), ensuring long-term
preservation (Preserve), enabling discovery and access (Access), and
managing final disposition (Disposition). This pervasive presence
underscores why metadata governance must be treated as an organizational
capability rather than a project-specific task, requiring dedicated
resources, trained personnel, and systematic processes that operate
continuously across all data activities.

\subsubsection{How our seven practices align to the data lifecycle
model}\label{how-our-seven-practices-align-to-the-data-lifecycle-model}

Our seven best practices map directly to established data lifecycle
models, demonstrating their broad applicability beyond salmon science.
The NOAA Data Lifecycle provides a widely recognized framework with six
sequential stages (Plan, Obtain, Process, Preserve, Access, Disposition)
and four cross-cutting elements (Document, Track and Monitor, Quality,
Security) (NOAA Data Governance Committee 2024). This alignment ensures
our practices are grounded in established federal data management
standards and can be readily adopted by other agencies and research
communities.

The mapping shown in Table~\ref{tbl-lifecycle} demonstrates how each
practice addresses specific lifecycle stages while the cross-cutting
elements ensure comprehensive data stewardship throughout the entire
lifecycle. For example, Practice 1 (Data Governance) spans the entire
lifecycle from planning through disposition, while Practice 4 (Shared
Data Models) primarily supports the Process and Preserve stages. This
systematic alignment with established frameworks enhances the
credibility and portability of our approach across different domains and
organizations.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.2689}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0672}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0756}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0840}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0924}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.0756}}
  >{\centering\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1176}}
  >{\raggedright\arraybackslash}p{(\linewidth - 14\tabcolsep) * \real{0.1681}}@{}}
\caption{Mapping of seven best practices to NOAA Data Lifecycle stages
and cross-cutting elements}\label{tbl-lifecycle}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Best Practice
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Plan
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Obtain
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Process
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Preserve
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Access
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Disposition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cross-cutting
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Best Practice
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Plan
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Obtain
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Process
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Preserve
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Access
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Disposition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cross-cutting
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1. Data Governance} & ● & ● & ● & ● & ● & ● & Document,
Quality \\
\textbf{2. Reuse Infrastructure} & ● & & & ● & ● & & Track and
Monitor \\
\textbf{3. Persistent Identifiers} & ● & & & ● & ● & ● & Document,
Track \\
\textbf{4. Shared Data Models} & ● & & ● & ● & & & Quality \\
\textbf{5. Accessible Storage} & & & ● & ● & ● & & Security, Quality \\
\textbf{6. Incentivize Sharing} & ● & & & ● & ● & & Track and Monitor \\
\textbf{7. Community Building} & ● & ● & ● & ● & ● & ● & Document,
Quality \\
\end{longtable}

\subsection{Conclusion}\label{conclusion}

Salmon biologists and data stewards across the globe have generated
extensive datasets on salmon abundance, environmental conditions, and
biological characteristics. When integrated, these data become valuable
assets, a fact powerfully demonstrated by studies such as (Peterman and
Dorner 2012). However, as noted by reports to the Cohen Commission
investigating the decline of Fraser River sockeye salmon (Marmorek et
al. 2011), these data are often incomplete, inconsistently collected,
and fragmented across organizations and jurisdictions. Integrating
across such diverse sources can be done but requires effort that is
often not accounted for in smaller-scale studies or perpetually
underestimated in long-term ongoing data projects. This fragmentation
hinders a deeper understanding of the drivers of change across salmon
life stages and regions, and limits the effectiveness of management
decisions, particularly concerning in the face of anthropic climate
change and biodiversity loss.

We argue that many persistent bottlenecks in salmon analysis and
management are fundamentally data problems, and that coordinated data
stewardship offers one of the most impactful, near-term opportunities
for improvement. By adopting shared best practices in data governance,
metadata standardization, persistent identification, infrastructure
reuse, and community co-development we can radically improve the
transparency, reusability, and interoperability of salmon data. A
coordinated, future-oriented data stewardship strategy can leverage the
full potential of data in science and management. The case study
presented in this paper---drawn from one of the Pacific Region's most
influential salmon survival syntheses (Peterman and Dorner
2012)---illustrates how technical and social data management gaps
directly obstructed efforts to answer pressing questions. If some of the
best practices we propose had been adopted by the data producers
substantial time and resources could have been saved. The case study
offers a clear and cautionary tale, as well as a hopeful roadmap.
Because data practices underpin nearly every analysis and decision,
strengthening data stewardship is one of the most effective ways salmon
programs can improve outcomes within their control.

The emergence of the data stewardship role (Plotkin 2014) represents one
of the most critical organizational shifts needed to realize this
vision. Historically, the work of managing, documenting, and maintaining
data has been diffuse and undervalued---often falling to biologists
without support, training, or recognition. As the volume and complexity
of scientific data grow, so too does the need for clearly defined data
stewardship responsibilities embedded within research teams and
organizations. Training biologists in the principles and practices of
data stewardship---while also supporting dedicated professionals who
specialize in this work---is essential to sustaining trustworthy,
reusable, and interoperable salmon data systems.

Realizing this vision requires concrete organizational commitments
organizations should formally appoint dedicated data stewards with clear
roles, responsibilities, and reporting structures. Agencies can adopt
centralized metadata repositories and establish compliance metrics to
track progress toward FAIR and CARE principles. Key implementation steps
include: (1) designating stewardship roles within existing
organizational structures, (2) investing in metadata management
infrastructure, (3) establishing data governance committees with
cross-organization representation, and (4) developing performance
indicators that measure data discoverability, interoperability, and
reuse. These organizational changes ensure that data stewardship becomes
embedded in organizational culture rather than remaining an ad hoc
responsibility.

The visionary future state is one where salmon researchers and
stewards---across agencies, Indigenous Nations, academic laboratories,
and community groups---can easily access and contribute to
well-documented, versioned, and machine-readable datasets. In this
future, field biologists, Indigenous guardians, modellers, and
policymakers interact with a living knowledge system---one that is
flexible, easy to implement, and rooted in principles of FAIRness and
Indigenous Data Sovereignty. Metadata standards, controlled
vocabularies, and shared governance frameworks are not afterthoughts but
integral to the culture of data collection and use. Scientists receive
credit for publishing high-quality data, and users trust the provenance
and structure of the datasets they rely on to make critical management
decisions.

Realizing this vision will require investment in both people and
systems. Key to this transformation is the emergence of the data steward
as a professional role: a hybrid expert who understands operational
field biology, information science, governance protocols, and community
needs. As highlighted by Roche et al. (2020), institutionalizing data
stewardship roles ensures long-term capacity for data governance,
quality control, and interoperability---functions that are often
neglected or left to informal actors. We must not only train new data
stewards but also support and upskill biologists to take on stewardship
responsibilities in collaborative, interdisciplinary settings. This is
essential to address the data technical debt of unmanaged data and to
modernize research practices in line with open science norms. By
embedding these practices into the everyday work of data generation,
documentation, publication, and reuse, we can move salmon science
decisively into the era of data-intensive discovery.

To do items:

\begin{itemize}
\tightlist
\item
  Discuss differences between data management plans, data governance
  charters, data sharing agreements
\item
  Incorporate ref to Streamnet Data Exchange Standards somehow
\item
  Add in figures
\item
  fill out appendix 1 more thoroughly
\item
  Refine Reorg the content in appendix 2 (traning roadmap) and decide if
  it Makes sense to put some of that content into a 3rd column in table
  1
\item
  Consider removing para 2 (Data Science adjancency) in Data Stewardship
  Defn
\item
  Consider re-working table to replace practical applications column
  with `next step'
\item
  Incorporate edits to Appendices from PSF folks
\item
  Incorporate edits from Gottfried Pestal
\end{itemize}

\subsubsection{Competing interests}\label{competing-interests}

\subsubsection{Acknowledgements}\label{acknowledgements}

\subsubsection{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-bull2022}
Bull, C D, S D Gregory, E Rivot, T F Sheehan, D Ensing, G Woodward, and
W Crozier. 2022. {``The Likely Suspects Framework: The Need for a Life
Cycle Approach for Managing Atlantic Salmon ({\emph{Salmo Salar}})
Stocks Across Multiple Scales.''} Edited by Wesley Flannery. \emph{ICES
Journal of Marine Science} 79 (5): 1445--56.
\url{https://doi.org/10.1093/icesjms/fsac099}.

\bibitem[\citeproctext]{ref-carroll2020}
Carroll, Stephanie Russo, Ibrahim Garba, Oscar L. Figueroa-Rodríguez,
Jarita Holbrook, Raymond Lovett, Simeon Materechera, Mark Parsons, et
al. 2020. {``The CARE Principles for Indigenous Data Governance.''}
\emph{Data Science Journal} 19.
\url{https://doi.org/10.5334/dsj-2020-043}.

\bibitem[\citeproctext]{ref-carroll2019}
Carroll, Stephanie Russo, Desi Rodriguez-Lonebear, and Andrew Martinez.
2019. {``Indigenous Data Governance: Strategies from United States
Native Nations.''} \emph{Data Science Journal} 18 (1): 31.
\url{https://doi.org/10.5334/dsj-2019-031}.

\bibitem[\citeproctext]{ref-diack2024}
Diack, Graeme, Tom Bird, Scott Akenhead, Jennifer Bayer, Deirdre Brophy,
Colin Bull, Elvira de Eyto, et al. 2024. {``Salmon Data Mobilization.''}
\emph{North Pacific Anadromous Fish Commission Bulletin}, December.
\url{https://doi.org/10.23849/npafcb7/x3rlpo23a}.

\bibitem[\citeproctext]{ref-eartheconomics2021}
Earth Economics. 2021. {``The Sociocultural Significance of Pacific
Salmon to Tribes and First Nations.''} Tacoma, Washington.
\url{https://www.psc.org/wp-content/uploads/wpfd/preview_files/The-Sociocultural-Significance-of-Salmon-to-Tribes-and-First-Nations(5da9942da9fb4fe0d77eb32bd6165e43).pdf}.

\bibitem[\citeproctext]{ref-noauthor_fraser_2025}
{``Fraser {River} {Panel} {Data} {Application}.''} 2025. Pacific Salmon
Commission. \url{https://psc1.shinyapps.io/PSC_In_Season_Fraser/}.

\bibitem[\citeproctext]{ref-groot1991}
Groot, Cornelis, and L. Margolis. 1991. \emph{Pacific Salmon Life
Histories}. UBC Press.
\url{https://books.google.ca/books?id=I_S0xCME0CYC}.

\bibitem[\citeproctext]{ref-inman2021}
Inman, Sarah, Janessa Esquible, Michael Jones, William Bechtol, and
Brendan Connors. 2021. {``Opportunities and Impediments for Use of Local
Data in the Management of Salmon Fisheries.''} \emph{Ecology and
Society} 26 (2). \url{https://doi.org/10.5751/ES-12117-260226}.

\bibitem[\citeproctext]{ref-jennings2023a}
Jennings, Lydia, Talia Anderson, Andrew Martinez, Rogena Sterling,
Dominique David Chavez, Ibrahim Garba, Maui Hudson, Nanibaa' A.
Garrison, and Stephanie Russo Carroll. 2023. {``Applying the {`}CARE
Principles for Indigenous Data Governance{'} to Ecology and Biodiversity
Research.''} \emph{Nature Ecology \& Evolution} 7 (10): 1547--51.
\url{https://doi.org/10.1038/s41559-023-02161-2}.

\bibitem[\citeproctext]{ref-johnson2024}
Johnson, Brett, and Tim van der Stap. 2024. {``Data Mobilization Through
the International Year of the Salmon Ocean Observing System.''}
\emph{North Pacific Anadromous Fish Commission Bulletin}, December.
\url{https://doi.org/10.23849/npafcb7/6a4ddpde4}.

\bibitem[\citeproctext]{ref-lindenmayer2012}
LINDENMAYER, DAVID B., GENE E. LIKENS, ALAN ANDERSEN, DAVID BOWMAN, C.
MICHAEL BULL, EMMA BURNS, CHRIS R. DICKMAN, et al. 2012. {``Value of
Long{-}Term Ecological Studies.''} \emph{Austral Ecology} 37 (7):
745--57. \url{https://doi.org/10.1111/j.1442-9993.2011.02351.x}.

\bibitem[\citeproctext]{ref-LocalContexts2025}
Local Contexts. 2025. {``Local Contexts: Grounding Indigenous Rights in
Data and Collections.''} {Local Contexts}. 2025.
\url{https://localcontexts.org/}.

\bibitem[\citeproctext]{ref-marmorek2011}
Marmorek, David, Darcy Pickard, Alexander Hall, Katherine Bryan, Liz
Martell, Clint Alexander, Katherine Wieckowski, Lorne Greig, and Carl
Schwarz. 2011. {``Cohen Commision Technical Report 6-Fraser River
Sockeye Salmon: Data Synthesis and Cumulative Impacts.''} Vancouver,
B.C. \url{http://www.cohencommission.ca/}.

\bibitem[\citeproctext]{ref-montenegro2019}
Montenegro, María. 2019. {``Subverting the Universality of Metadata
Standards.''} \emph{Journal of Documentation} 75 (4): 731--49.
\url{https://doi.org/10.1108/jd-08-2018-0124}.

\bibitem[\citeproctext]{ref-environm2007}
NOAA. 2007. \emph{Environmental Data Management at NOAA}. National
Academies Press. \url{https://doi.org/10.17226/12017}.

\bibitem[\citeproctext]{ref-noaadatagovernancecommittee2024}
NOAA Data Governance Committee. 2024. {``Management of NOAA Data and
Information, Data Management Handbook,''} January.
\url{https://www.noaa.gov/sites/default/files/2025-03/NAO_212-15B_-_Data_Management_Handbook.pdf}.

\bibitem[\citeproctext]{ref-peng2018}
Peng, Ge, Jeffrey L. Privette, Curt Tilmes, Sky Bristol, Tom Maycock,
John J. Bates, Scott Hausman, Otis Brown, and Edward J. Kearns. 2018.
{``A Conceptual Enterprise Framework for Managing Scientific Data
Stewardship.''} \emph{Data Science Journal} 17.
\url{https://doi.org/10.5334/dsj-2018-015}.

\bibitem[\citeproctext]{ref-peterman2012}
Peterman, Randall~M., and Brigitte Dorner. 2012. {``A Widespread
Decrease in Productivity of Sockeye Salmon ({\emph{Oncorhynchus Nerka}})
Populations in Western North America.''} Edited by Jordan S. Rosenfeld.
\emph{Canadian Journal of Fisheries and Aquatic Sciences} 69 (8):
1255--60. \url{https://doi.org/10.1139/f2012-063}.

\bibitem[\citeproctext]{ref-plotkin2014}
Plotkin, David. 2014. \emph{Data Stewardship}. Elsevier.
\url{https://doi.org/10.1016/c2012-0-07057-3}.

\bibitem[\citeproctext]{ref-roche2020}
Roche, Dominique G., Monica Granados, Claire C. Austin, Scott Wilson,
Gregory M. Mitchell, Paul A. Smith, Steven J. Cooke, and Joseph R.
Bennett. 2020. {``Open Government Data and Environmental Science: A
Federal Canadian Perspective.''} Edited by Tanzy Love. \emph{FACETS} 5
(1): 942--62. \url{https://doi.org/10.1139/facets-2020-0008}.

\bibitem[\citeproctext]{ref-volk2014}
Volk, Carol J., Yasmin Lucero, and Katie Barnas. 2014. {``Why Is Data
Sharing in Collaborative Natural Resource Efforts so Hard and What Can
We Do to Improve It?''} \emph{Environmental Management} 53 (5): 883--93.
\url{https://doi.org/10.1007/s00267-014-0258-2}.

\bibitem[\citeproctext]{ref-ward_surveyjoin_2025}
Ward, Eric J, Philina A English, Christopher N Rooper, Bridget E
Ferriss, Curt E Whitmire, Chantel R Wetzel, Lewis Ak Barnett, et al.
2025. {```Surveyjoin`: {A} {Standardized} {Database} of {Fisheries}
{Bottom} {Trawl} {Surveys} in the {Northeast} {Pacific} {Ocean}.''}
\url{https://doi.org/10.1101/2025.03.14.643022}.

\bibitem[\citeproctext]{ref-ween2013}
Ween, Gro, and Benedict Colombi. 2013. {``Two Rivers: The Politics of
Wild Salmon, Indigenous Rights and Natural Resource Management.''}
\emph{Sustainability} 5 (2): 478--95.
\url{https://doi.org/10.3390/su5020478}.

\bibitem[\citeproctext]{ref-wilkinson2016}
Wilkinson, Mark D., Michel Dumontier, IJsbrand Jan Aalbersberg,
Gabrielle Appleton, Myles Axton, Arie Baak, Niklas Blomberg, et al.
2016. {``The FAIR Guiding Principles for Scientific Data Management and
Stewardship.''} \emph{Scientific Data} 3 (1): 160018.
\url{https://doi.org/10.1038/sdata.2016.18}.

\end{CSLReferences}

\section{Appendices}\label{appendices}

\section{}\label{section}

\subsection{Appendix 1. Real-world Example Applications of the Best
Practices}\label{appendix-1.-real-world-example-applications-of-the-best-practices}

Here we provide detailed descriptions of the seven best practices for
salmon data stewardship, along with practical applications and
real-world examples. This is not an exhaustive list, but rather a
starting point for salmon biologists and data stewards to implement
effective data stewardship practices in their work based on examples
from the salmon research and management community.

\paragraph{\texorpdfstring{\textbf{1. Make Data Governance Explicit to
Support Trust and
Reuse}}{1. Make Data Governance Explicit to Support Trust and Reuse}}\label{make-data-governance-explicit-to-support-trust-and-reuse}

Clear governance defines roles, responsibilities, and procedures
ensuring data quality, long-term maintenance, accountability, and
compliance with community principles such as FAIR and CARE. Effective
governance fosters trust, facilitates data sharing, reduces ambiguity
regarding decision making, and is critical for coordinating both
technical and sociocultural aspects of data stewardship.

In collaborative international or multi-organizational settings,
establishing governance at the outset of a project is crucial for
aligning diverse groups, including biologists, data managers, Indigenous
communities, policymakers, and other participants. Early governance
planning should establish clear, collaborative frameworks that respect
each group's expertise and needs from the beginning.~

\subparagraph{Practical Applications:}\label{practical-applications}

1.1 Document roles and responsibilities clearly at project start using a
Project or Data Product Governance Charter and structured frameworks
(e.g., DACI or RACI charts) that relate to organizational data policies.

\begin{itemize}
\item
  \href{https://raw.githubusercontent.com/Br-Johnson/sdm-paper/refs/heads/main/examples/Cal_DMP.pdf}{Example
  of a Data Management Plan from the California Department of Water
  Resources}
\item
  Data Management Plan Templates from
  \href{https://dmptool.org/public_templates?page=ALL}{DMPTool}, and
  \href{https://www.noaa.gov/sites/default/files/2025-03/NAO_212-15B_-_Data_Management_Handbook.pdf}{NOAA
  Data Management Handbook}
\end{itemize}

1.2 Integrate CARE principles to ensure ethical governance and respect
Indigenous data rights.

\begin{itemize}
\tightlist
\item
  Northwest Indian Fisheries Commission use password protected website
  to host all the WDFW and tribal data in a one-stop shopping website
  for co-managers to pull data they need for decision-making process.
  \url{https://fisheriesservices.nwifc.org/}
\end{itemize}

1.3 Create a governance or oversight committee for regular data practice
reviews and decision making regarding data structures, timelines, data
sharing agreements and interoperability protocols

\begin{itemize}
\tightlist
\item
  Pacific Salmon Commission has formed a Technical Committee on Data
  Sharing including both US and Canadian data contributors.
  \url{https://www.psc.org/membership-lists/}
\end{itemize}

\paragraph{\texorpdfstring{\textbf{2. Reuse Proven Infrastructure to
Save Time and Increase
Interoperability}}{2. Reuse Proven Infrastructure to Save Time and Increase Interoperability}}\label{reuse-proven-infrastructure-to-save-time-and-increase-interoperability}

Building custom solutions should be avoided where possible. Maximizing
existing platforms and technologies reduces costs, accelerates
implementation, and increases data interoperability. Building modular,
interoperable systems grounded in proven technologies ensures
sustainable long-term stewardship.

\subparagraph{Practical Applications:}\label{practical-applications-1}

2.1 \href{https://doi.org/10.14286/duc6mu}{Use the Ocean Biodiversity
Information System or the Global Biodiversity Information Facility to
standardize and host your data}

2.2 Use free data catalogue services such as the Knowledge Network for
Biocomplexity (KNB) or Zenodo

\begin{itemize}
\tightlist
\item
  The Pacific Salmon Foundation's
  \href{https://doi.org/10.5281/zenodo.14194638}{spawner surveys dataset
  on Zenodo} (Carturan and Peacock 2025) received more views within
  weeks than a
  \href{https://data.salmonwatersheds.ca/result?datasetid=2}{analogous
  dataset in the Salmon Data Library} (Pacific Salmon Foundation 2025)
  did over several years, illustrating that leveraging established
  public data infrastructures, rather than developing
  institution-specific ones, can substantially increase discoverability.
\end{itemize}

\paragraph{\texorpdfstring{\textbf{3. Make Data, People, Projects, and
Outputs Discoverable, Linked and Citable with
PIDs}}{3. Make Data, People, Projects, and Outputs Discoverable, Linked and Citable with PIDs}}\label{make-data-people-projects-and-outputs-discoverable-linked-and-citable-with-pids}

Persistent identifiers (PIDs), including Digital Object Identifiers
(DOI) are essential for tracking the provenance and reuse of data, and
linking data, protocols, organizations and people. They allow for
consistent referencing, integration across systems, and automated credit
via data citation.

\subparagraph{Practical Applications:}\label{practical-applications-2}

3.1 Encourage researchers to register for an
\href{https://orcid.org/}{Open Researcher and Contributor ID (ORCID)}
and include ORCIDs in metadata records and submission forms

3.2 Register your organization with the Research Organization Registry
(ROR) and use ROR IDs to identify institutions involved in salmon
science.

\begin{itemize}
\tightlist
\item
  Several salmon data holding institutions are already
  \href{https://ror.org/search?page=1&query=salmon}{registered with
  ROR}. As a result, those organizations can track and demonstrate their
  scholarly impact from data publications:
  \href{https://commons.datacite.org/ror.org/04901nj56}{DataCite
  Commons: Pacific Salmon Foundation}
\end{itemize}

3.3 Assign DOIs to data packages, protocols, and reports using DataCite.
Maintain version history for all metadata records and document the
provenance of metadata creation, updates, and quality control processes
to ensure accountability and traceability.

\begin{itemize}
\tightlist
\item
  The North Pacific Anadromous FIsh Commission (NPAFC) assigns DOIs to
  IYS-related data packages which are served by a CKAN catalogue at
  \url{https://data.npafc.org}. The Commission also assigns DOIs to
  NPAFC Technical Reports and Bulletins.
\end{itemize}

3.4 Embed DOIs in dashboards, figures, and metadata so they persist in
derivative products.

\paragraph{\texorpdfstring{\textbf{4. Use Shared Data Models,
Vocabularies and Metadata to Enable
Integration}}{4. Use Shared Data Models, Vocabularies and Metadata to Enable Integration}}\label{use-shared-data-models-vocabularies-and-metadata-to-enable-integration}

Standardizing metadata and terminology ensures data can be interpreted
correctly and integrated across systems. Controlled vocabularies,
community ontologies, and structured metadata schemas allow data to
retain its full semantic meaning.

\subparagraph{Practical Applications:}\label{practical-applications-3}

4.1 Configure data catalogues and metadata intake tools to accept
internationally recognized metadata schemas such as ISO 19115,
Ecological Metadata Language (EML), or DataCite. Implement automated
validation against schema requirements and manual review processes to
ensure metadata completeness, accuracy, and consistency before
publication.

\begin{itemize}
\tightlist
\item
  The Pacific Salmon Foundation's data portal asks contributors to
  provide metadata in ISO 19115 or other standard formats.
  marinedata.psf.ca, ensuring consistent metadata structure
\item
  The NPAFC uses ISO 19115 metadata standard in their data catalogue
  https://data.npafc.org
\end{itemize}

4.2 Model datasets and databases using the
\href{https://dwc.tdwg.org/}{Darwin Core Standard}

\begin{itemize}
\item
  The Hakai Institute Juvenile Salmon Program publishes their data to
  OBIS using Darwin Core:
  \href{https://www.gbif.org/dataset/72de3af4-1572-4f2d-8006-2bfa2007065c}{Hakai
  Institute Juvenile Salmon Program}
\item
  The International Year of the Salmon High Seas Expeditions data
  mobilization efforts {[}Johnson and Stap (2024){]} published their
  data to OBIS: \url{https://www.gbif.org/dataset/search?project_id=IYS}
\end{itemize}

4.3 Re-use or publish data terms that are shared online using a
persistent identifier in a controlled vocabulary or ontology

\begin{itemize}
\item
  DFO Salmon Ontology\ldots{}
\item
  State of Alaska Salmon and People\ldots{}
\item
  Measurement Types in OBIS\ldots{}
\item
  WDFW has definitions of all hatchery escapement data.
  \href{https://wdfw.wa.gov/fishing/management/hatcheries/escapement\#definitions}{Hatchery
  escapement reports \textbar{} Washington Department of Fish \&
  Wildlife}
\item
  Fish Passage Counts has defined metadata that can be used across OFDW
  and WDFW.
  \url{https://www.fpc.org/111_sharedfiles/ColumbiaRiverBasinAdultFishPassageCountsMetadata.pdf}
\end{itemize}

\paragraph{\texorpdfstring{\textbf{Best Practice 5: Store and Analyze
Data in Ways That Others Can Easily Access, Use, and
Trust}}{Best Practice 5: Store and Analyze Data in Ways That Others Can Easily Access, Use, and Trust}}\label{best-practice-5-store-and-analyze-data-in-ways-that-others-can-easily-access-use-and-trust}

Making data easily accessible promotes its use in research and
management, enabling seamless integration with tools and applications.
Ensuring accessible, persistent data storage requires more than just
file hosting. Data should be structured, accessible via API, and stored
in repositories that support long-term preservation.

\subparagraph{Practical Applications:}\label{practical-applications-4}

5.1 Provide Direct Data Access via Application Programming Interfaces
(APIs) using tools such as FastAPI, Flask, or Django REST Framework that
allows users to access, filter, and retrieve data programmatically,
facilitating automation and integration into analytical tools and
decision-support systems

\begin{itemize}
\tightlist
\item
  The Pacific States Marine Fisheries Commission make's their
  \href{https://www.psmfc.org/program/pit-tag-information-systems-ptagis}{PIT
  Tag Information System} data accessible via the
  \href{https://api.ptagis.org/index.html\#:~:text=PTAGIS\%20API\%20Gets\%20a\%20list,PTAGIS\%20and\%20currently\%20contributing\%20data}{PTAGIS
  API}
\end{itemize}

5.2 Archive data in certified long-term, domain-specific repositories
such as the Global Biodiversity Information Facility, the Federated
Research Data Repository (FRDR), or NOAA's NCEI, USGS ScienceBase, or
EMODnet

\begin{itemize}
\tightlist
\item
  TODO
\end{itemize}

5.3 Leverage the integration between GitHub and Zenodo to automate
archiving and DOI assignment, ensuring long-term data preservation.

\paragraph{\texorpdfstring{\textbf{6. Incentivize and Track Data Sharing
and
Reuse}}{6. Incentivize and Track Data Sharing and Reuse}}\label{incentivize-and-track-data-sharing-and-reuse}

The currency of research lies in recognition---credit, citations, and
opportunities for collaboration or co-authorship. Promoting data sharing
requires both cultural and technical infrastructure. The cultural
infrastructure requires a shift towards viewing data publication as
equal in importance to article publication. The infrastructure put in
place needs to support the process of generating citation records that
give credit to all First Nations, Tribes, agencies, and organizations.
By recognizing contributions, tracking reuse, and supporting citation,
data stewards can create a system where sharing is rewarded.

\subparagraph{Practical Applications:}\label{practical-applications-5}

6.1 License data for reuse using liberal licenses

\begin{itemize}
\tightlist
\item
  All data accessible through the \href{https://data.npafc.org}{NPAFC
  data catalogue} is licenced as~
  \href{https://creativecommons.org/licenses/by/4.0/deed.en}{Creative
  Commons Attribution 4.0 International}
\end{itemize}

6.2 Provide recommended citation text and visible credit fields in
metadata

6.3 Create summary dashboards that highlight reuse using
\href{https://www.countermetrics.org/}{COUNTER Code of Practice
compliant metrics} to track dataset views/downloads and the DataCite
APIs

6.4 Ensure that datasets are properly cited in journal articles using in
text citations and the recommended citation in the articles list of
references, not just in a Data Availability statement

\begin{itemize}
\tightlist
\item
  In late 2024, the NPAFC began citing data sets using in-text citations
  and the recommended citation in the list of references with the
  publication of NPAFC Bulletin 7 titled, \emph{Highlights of the 2022
  International Year of the Salmon Pan--Pacific Winter Expedition}.
\end{itemize}

6.5 Promote the view that well documented data publications are primary
research outputs and are significant contributions to the field

\paragraph{\texorpdfstring{\textbf{7. Build Community Through
Co-Development and Mutual
Benefit}}{7. Build Community Through Co-Development and Mutual Benefit}}\label{build-community-through-co-development-and-mutual-benefit}

Creating an infrastructure that standardizes and provides cross-border
and cross-ecosystem data integration is only effective if there's
community engagement. Standards and tools must be co-developed with
their intended users using user-centred design principles (citation
required) to be effective. Engaging biologists, Indigenous stewards, and
data managers ensures relevance, usability, and long-term participation.

\subparagraph{Practical Applications:}\label{practical-applications-6}

7.1 Participate in salmon data focused communities such as the
\href{https://www.rd-alliance.org/groups/salmon-research-and-monitoring-ig/}{Research
Data Alliance's Salmon Research and Monitoring Interest Group}~

7.2 Run participatory workshops for metadata mapping and vocabulary
alignment

\begin{itemize}
\tightlist
\item
  American Fisheries Society 2025 WA-BC Chapter Annual Meeting workshop.
  \href{https://pnamp.org/event/workshop-fishing-for-clarity-knowledge-modeling-to-support-cross-organizational-collaboration-and-data-sharing-about-salmon-escapement/}{'Fishing
  for Clarity: Knowledge Modelling to Support Cross-organizational
  Collaboration and Data Sharing about Salmon Escapement}
\end{itemize}

7.3 Support and follow through on Community Engaged Research
(e.g.~\href{https://salmonprize.com}{The Salmon Prize Project}) that
provides tangible value to the communities in which research or
monitoring was conducted.

\textsubscript{Source:
\href{https://br-johnson.github.io/sdm-paper/appendices/appendix-1-real-world-examples-preview.html\#f74d3bc4-54eb-4715-971f-d77b337c50f9}{Appendix
1. Real-world Example Applications of the Best Practices}}

\section{}\label{section-1}

\subsection{Appendix 2: Training Roadmap for Salmon Biologists
Transitioning to Data
Stewardship}\label{appendix-2-training-roadmap-for-salmon-biologists-transitioning-to-data-stewardship}

This roadmap outlines essential topics, resources, and learning
materials salmon biologists should engage with to effectively transition
into data stewardship roles. The roadmap follows a structured
progression similar to \href{https://roadmap.sh}{roadmap.sh}.

\subsubsection{1. Foundations of Data
Stewardship}\label{foundations-of-data-stewardship}

\begin{itemize}
\tightlist
\item
  \textbf{Principles:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://www.go-fair.org/fair-principles/}{FAIR Data
    Principles}
  \item
    \href{https://www.gida-global.org/care}{CARE Principles for
    Indigenous Data Governance}
  \end{itemize}
\item
  \textbf{Seminal Papers:}

  \begin{itemize}
  \tightlist
  \item
    Wilkinson et al.~2016
    \href{https://doi.org/10.1038/sdata.2016.18}{FAIR Guiding
    Principles}
  \item
    Carroll et al.~2019
    \href{https://doi.org/10.5334/dsj-2019-031}{Indigenous Data
    Governance}
  \end{itemize}
\item
  \textbf{Courses and Tutorials:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://www.go-fair.org/resources/videos/fair-principles-explained/}{FAIR
    Principles Explained (GO-FAIR)}
  \end{itemize}
\end{itemize}

\subsubsection{2. Data Management \&
Governance}\label{data-management-governance}

\begin{itemize}
\tightlist
\item
  \textbf{Seminal Papers and Reports:}

  \begin{itemize}
  \tightlist
  \item
    Plotkin, 2014 \href{https://doi.org/10.1016/c2012-0-07057-3}{Data
    Stewardship}
  \item
    NOAA, 2007 \href{https://doi.org/10.17226/12017}{Environmental Data
    Management}
  \end{itemize}
\item
  \textbf{Practical Tools:}

  \begin{itemize}
  \tightlist
  \item
    Data Management Plan Templates (\href{https://dmptool.org}{DMPTool})
  \item
    DACI and RACI Frameworks
    (\href{https://www.atlassian.com/team-playbook/plays/daci}{Atlassian
    DACI Guide})
  \end{itemize}
\end{itemize}

\subsubsection{3. Metadata Standards and
Ontologies}\label{metadata-standards-and-ontologies}

\begin{itemize}
\tightlist
\item
  \textbf{Standards to Master:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://www.iso.org/standard/53798.html}{ISO 19115 Metadata
    Standard}
  \item
    \href{https://eml.ecoinformatics.org/}{Ecological Metadata Language
    (EML)}
  \item
    \href{https://dwc.tdwg.org/}{Darwin Core Standard}
  \end{itemize}
\item
  \textbf{Case Studies \& Examples:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://marinedata.psf.ca}{Pacific Salmon Foundation Metadata
    Standards}
  \item
    \href{https://obis.org/dataset/81a9b931-8f47-49c4-94ff-0e46d7e9f1c9}{Hakai
    Institute Juvenile Salmon Program}
  \end{itemize}
\end{itemize}

\subsubsection{4. Controlled Vocabularies \& Persistent Identifiers
(PIDs)}\label{controlled-vocabularies-persistent-identifiers-pids}

\begin{itemize}
\tightlist
\item
  \textbf{PIDs to Implement:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://datacite.org}{DOIs via DataCite}
  \item
    \href{https://orcid.org}{ORCID IDs}
  \item
    \href{https://ror.org}{Research Organization Registry (ROR)}
  \end{itemize}
\item
  \textbf{Practical Guides:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://blog.datacite.org/}{Introduction to PIDs (DataCite
    Blog)}
  \end{itemize}
\end{itemize}

\subsubsection{5. Data Integration \&
Interoperability}\label{data-integration-interoperability}

\begin{itemize}
\tightlist
\item
  \textbf{Seminal Papers:}

  \begin{itemize}
  \tightlist
  \item
    Johnson \& Stap, 2024
    \href{https://doi.org/10.23849/npafcb7/6a4ddpde4}{Salmon Ocean
    Observing System}
  \item
    Bull et al.~2022
    \href{https://doi.org/10.1093/icesjms/fsac099}{Likely Suspects
    Framework}
  \end{itemize}
\item
  \textbf{Technical Skills \& Tools:}

  \begin{itemize}
  \tightlist
  \item
    APIs with \href{https://fastapi.tiangolo.com}{FastAPI},
    \href{https://flask.palletsprojects.com}{Flask},
    \href{https://www.django-rest-framework.org}{Django REST Framework}
  \item
    \href{https://guides.github.com/activities/citable-code/}{Zenodo-GitHub
    Integration}
  \end{itemize}
\end{itemize}

\subsubsection{6. Data Sharing, Citation \&
Metrics}\label{data-sharing-citation-metrics}

\begin{itemize}
\tightlist
\item
  \textbf{Best Practices:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons
    Attribution 4.0 International License}
  \end{itemize}
\item
  \textbf{Tracking \& Metrics:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://www.projectcounter.org/}{COUNTER Metrics}
  \item
    \href{https://datacite.org/service.html}{DataCite API}
  \end{itemize}
\end{itemize}

\subsubsection{7. Community Engagement \&
Co-Development}\label{community-engagement-co-development}

\begin{itemize}
\tightlist
\item
  \textbf{Communities \& Groups:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://rd-alliance.org/groups/salmon-data-management}{Research
    Data Alliance Salmon Research Group}
  \end{itemize}
\item
  \textbf{Approaches \& Frameworks:}

  \begin{itemize}
  \tightlist
  \item
    User-centered Design
    (\href{https://www.interaction-design.org/literature/topics/user-centered-design}{Interaction
    Design Foundation})
  \item
    Community Engaged Research
    (\href{https://www.uvic.ca/research/assets/docs/Community-Engaged-Research.pdf}{University
    of Victoria Guide})
  \end{itemize}
\end{itemize}

\subsubsection{Additional Resources}\label{additional-resources}

\begin{itemize}
\tightlist
\item
  \textbf{Free Courses:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://www.fosteropenscience.eu/toolkit}{Introduction to Open
    Science (FOSTER)}
  \item
    \href{https://www.coursera.org/learn/data-management}{Research Data
    Management and Sharing (Coursera)}
  \end{itemize}
\item
  \textbf{Blogs \& Websites:}

  \begin{itemize}
  \tightlist
  \item
    \href{https://blog.okfn.org/}{Open Knowledge Foundation Blog}
  \item
    \href{https://www.dataone.org/resources/}{DataONE Data Management
    Resources}
  \end{itemize}
\end{itemize}

This roadmap serves as a structured guide to equip salmon biologists
with the practical and theoretical skills required to excel in data
stewardship roles.

\textsubscript{Source:
\href{https://br-johnson.github.io/sdm-paper/appendices/appendix-2-training-roadmap-preview.html\#718196ae-8a7f-4f9d-94a8-00b0045c1356}{Appendix
2: Training Roadmap for Salmon Biologists Transitioning to Data
Stewardship}}

\section{}\label{section-2}

\subsection{Appendix 3: Getting Started
Checklist}\label{appendix-3-getting-started-checklist}

\section{Getting Started Checklist for Salmon Data
Stewardship}\label{getting-started-checklist-for-salmon-data-stewardship}

Use this practical checklist to assess how well your project, program,
or organization aligns to the seven Best Practices. Start at the Project
level, then scale to Program and Organization. Check off items you've
completed and note gaps to prioritize.

Tip: For each item, capture a link to the living source (e.g.,
repository, shared drive, policy page) and the responsible owner.

\subsection{Practice 1 --- Make Data Governance
Explicit}\label{practice-1-make-data-governance-explicit}

\paragraph{Project}\label{project}

\begin{verbatim}
- [ ] Do you have a Data Management Plan (DMP) covering scope, sensitive data, retention, and sharing? (link)
- [ ] Is there a RACI (Responsible, Accountable, Consulted, Informed) table for key tasks? (owner)
- [ ] Are Indigenous knoweledge holders or community members involved in the project?
- [ ] Are Indigenous Data Sovereignty (IDS) requirements identified and documented (who to consult, approvals needed)?
- [ ] Is a data product charter written for each dataset or analysis product with purpose, audience, quality thresholds, release plan?
\end{verbatim}

\paragraph{Program}\label{program}

\begin{verbatim}
- [ ] Are DMP and charter templates standardized across projects and stored centrally?
- [ ] Are role definitions for Data Steward, Product Owner, and Maintainer explicit and assigned for priority datasets?
- [ ] Is this 'community-engaged' research that provides tangible benefit to communities?
- [ ] Are data sharing agreements/MOUs and ethical review pathways documented and reusable?
\end{verbatim}

\paragraph{Organization}\label{organization}

\begin{verbatim}
- [ ] Does a governance policy exist that sets minimum requirements for DMPs, RACI, retention, IDS, and release reviews?
- [ ] Is there a standing review forum (e.g., monthly data governance check‑in) and a registry of governed data products?
\end{verbatim}

Evidence to collect: DMP link, data product charter(s), RACI, IDS
guidance, sharing agreements registry.

\subsection{Practice 2 --- Reuse Proven
Infrastructure}\label{practice-2-reuse-proven-infrastructure}

\paragraph{Project}\label{project-1}

\begin{verbatim}
- [ ] Have you researched the existing data sharing infrastructure and data storage options specific to your data and context? Eg. Ocean Biodiversity Information System, Global Biodiversity Information Facility, Knowledge Network for Biocomplexity, Zenodo, Dataverse, etc.
- [ ] Is your code in version control (e.g., Git) with an issue tracker and releases?
- [ ] Are you using an approved repository or data store rather than creating a new silo? (where)
- [ ] Do you use existing organization authentication/authorization and backup processes?
\end{verbatim}

\paragraph{Program}\label{program-1}

\begin{verbatim}
- [ ] Is there a preferred stack list (storage, metadata catalog, workflow runner, packaging, container base images)?
- [ ] Do projects consistently deposit finalized data in approved repositories with clear intake criteria?
\end{verbatim}

\paragraph{Organization}\label{organization-1}

\begin{verbatim}
- [ ] Are enterprise services available and documented (data lake, object store, catalog/portal, archival repository)?
- [ ] Is there a deprecation pathway for legacy systems and a migration plan for priority datasets?
\end{verbatim}

Evidence to collect: repository URLs, infrastructure inventory, intake
criteria, backup/DR documentation.

\subsection{Practice 3 --- Use Persistent Identifiers (PIDs) for People,
Projects, Data, and
Methods}\label{practice-3-use-persistent-identifiers-pids-for-people-projects-data-and-methods}

\paragraph{Project}\label{project-2}

\begin{verbatim}
- [ ] Do all contributors have ORCID IDs recorded in metadata?
- [ ] Does the project have a resolvable PID (e.g., DOI for a project page or protocol, internal project ID)?
- [ ] Are datasets assigned DOIs (or other PIDs) at publication, and are versions tracked?
- [ ] Are methods/protocols published and citable (e.g., protocol DOI) and linked from dataset metadata?
\end{verbatim}

\paragraph{Program}\label{program-2}

\begin{verbatim}
- [ ] Is there guidance on when to mint PIDs, by whom, and where they resolve?
- [ ] Are projects linked to organizational identifiers (e.g., ROR for institutions) in metadata?
\end{verbatim}

\paragraph{Organization}\label{organization-2}

\begin{verbatim}
- [ ] Is there a PID policy and a provider/registrar configured (e.g., DataCite) with a documented workflow?
- [ ] Are PID linkages automated in the catalog (people ↔ projects ↔ datasets ↔ publications)?
\end{verbatim}

Evidence to collect: ORCID list, PID policy, DOI records, resolver links
in the catalog.

\subsection{Practice 4 --- Shared Data Models, Vocabularies, and
Metadata}\label{practice-4-shared-data-models-vocabularies-and-metadata}

\paragraph{Project}\label{project-3}

\begin{verbatim}
- [ ] Which metadata profile is used (e.g., ISO 19115, EML)? Is the minimum profile complete and machine‑readable?
- [ ] Are core entities modeled consistently (stock/population IDs, locations, temporal coverage, age/brood year semantics)?
- [ ] Are controlled vocabularies/code lists applied for key fields (e.g., species codes, gear, life stage, age type)?
- [ ] Is a data dictionary included with definitions, units, allowed values, and provenance for each variable?
\end{verbatim}

\paragraph{Program}\label{program-3}

\begin{verbatim}
- [ ] Do projects use a shared schema and code lists across datasets to enable easy joins?
- [ ] Are validation checks in CI (schema validation, vocab checks) standardized across repositories?
\end{verbatim}

\paragraph{Organization}\label{organization-3}

\begin{verbatim}
- [ ] Is there an endorsed salmon domain profile and shared code lists with owners and change control?
- [ ] Are mappings to external standards maintained (e.g., taxonomic, geospatial, hydrological registries)?
\end{verbatim}

Evidence to collect: metadata profiles, data dictionary, code lists,
schema validators, mapping documentation.

\subsection{Practice 5 --- Store and Analyze Data for Easy Access, Use,
and
Trust}\label{practice-5-store-and-analyze-data-for-easy-access-use-and-trust}

\paragraph{Project}\label{project-4}

\begin{verbatim}
- [ ] Is raw data immutable and separated from processed/analysis outputs?
- [ ] Is there a fully reproducible workflow (scripts/notebooks + environment + parameters) that runs end‑to‑end?
- [ ] Is the computational environment captured (lockfile/conda env, container image) and versioned?
- [ ] Are QA/QC checks automated with logs and thresholds documented?
- [ ] Are access controls and sensitive data handling documented and implemented?
\end{verbatim}

\paragraph{Program}\label{program-4}

\begin{verbatim}
- [ ] Do projects follow a common repo layout and release process (tags, changelog, signed artifacts)?
- [ ] Are standard storage classes, lifecycle policies, and archival rules applied?
\end{verbatim}

\paragraph{Organization}\label{organization-4}

\begin{verbatim}
- [ ] Are security, backup/retention, and audit requirements defined and routinely verified?
- [ ] Is there a trusted long‑term archive with fixity checking and preservation metadata?
\end{verbatim}

Evidence to collect: workflow definition, environment files, container
references, QA/QC reports, storage/backup settings.

\subsection{Practice 6 --- Incentivize and Track Sharing and
Reuse}\label{practice-6-incentivize-and-track-sharing-and-reuse}

\paragraph{Project}\label{project-5}

\begin{verbatim}
- [ ] Is a clear citation and license statement included in metadata and README?
- [ ] Are reuse metrics collected (downloads, citations, API hits) and reviewed?
- [ ] Do release notes document what changed and implications for users?
\end{verbatim}

\paragraph{Program}\label{program-5}

\begin{verbatim}
- [ ] Are common metrics dashboards available for priority datasets and updated automatically?
- [ ] Are data citations tracked in assessments, reports, and staff evaluations?
\end{verbatim}

\paragraph{Organization}\label{organization-5}

\begin{verbatim}
- [ ] Do policies require citation guidance and permissive, appropriate licensing where possible?
- [ ] Are automated reports of reuse (e.g., via DOI provider APIs) delivered to product owners and leadership?
\end{verbatim}

Evidence to collect: LICENSE, CITATION, reuse dashboard link, policy
excerpts, sample citations in reports.

\subsection{Practice 7 --- Build Community Through Co‑Development and
Mutual
Benefit}\label{practice-7-build-community-through-codevelopment-and-mutual-benefit}

\paragraph{Project}\label{project-6}

\begin{verbatim}
- [ ] Are stakeholders identified, including First Nations/Tribes/Indigenous partners, and engagement needs documented?
- [ ] Have you held at least one co‑design session to validate user needs and success criteria?
- [ ] Is there an open feedback channel (issues form, contact) and a published roadmap?
\end{verbatim}

\paragraph{Program}\label{program-6}

\begin{verbatim}
- [ ] Do cross‑project working groups exist for models, vocabularies, and tooling with regular cadence and notes?
- [ ] Are community contributions recognized (authorship, acknowledgements, meeting time, funding)?
\end{verbatim}

\paragraph{Organization}\label{organization-6}

\begin{verbatim}
- [ ] Is there an endorsed governance body or community of practice with decision records?
- [ ] Are procurement/funding mechanisms available to support shared components and Indigenous partnerships?
\end{verbatim}

Evidence to collect: stakeholder map, engagement records, roadmap,
working group notes, decision log.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{Quick Start: 30/60/90‑Day
Plan}\label{quick-start-306090day-plan}

\begin{itemize}
\tightlist
\item
  First 30 days

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Create/standardize DMP + RACI; draft data product charters for top
    1--2 datasets.
  \item[$\square$]
    Move code to version control; document repo structure; capture
    environment file.
  \item[$\square$]
    Choose metadata profile and draft a minimal data dictionary; list
    code lists in use.
  \end{itemize}
\item
  By 60 days

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Mint/plan PIDs (project page/protocols), add ORCIDs to metadata,
    prepare DOI for first dataset.
  \item[$\square$]
    Add schema + vocab validation to CI; separate raw/processed;
    automate QA/QC checks.
  \item[$\square$]
    Stand up a reuse dashboard or basic metrics capture; add
    citation/license to README and metadata.
  \end{itemize}
\item
  By 90 days

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Publish first governed release to approved repository with DOI and
    complete metadata.
  \item[$\square$]
    Formalize cross‑project working group and change control for
    vocabularies.
  \item[$\square$]
    Document archival/retention path and verify backups; schedule
    governance reviews.
  \end{itemize}
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{Minimal Artifacts Checklist (Project
Level)}\label{minimal-artifacts-checklist-project-level}

\begin{itemize}
\tightlist
\item[$\square$]
  DMP (with IDS considerations) and RACI
\item[$\square$]
  Data product charter(s) for priority dataset(s)
\item[$\square$]
  Versioned repository with releases and changelog
\item[$\square$]
  Metadata profile file + data dictionary + code lists
\item[$\square$]
  Reproducible workflow + environment file or container
\item[$\square$]
  QA/QC checks and results log
\item[$\square$]
  Citation and license statements
\item[$\square$]
  Plan for PIDs (ORCID list, dataset/protocol DOIs)
\item[$\square$]
  Evidence of stakeholder engagement and roadmap
\end{itemize}

Maintain this list as a living issue in your repository and review
quarterly.

\textsubscript{Source:
\href{https://br-johnson.github.io/sdm-paper/appendices/appendix-3-getting-started-checklist-preview.html\#ed8748e7-1461-446c-a601-f2e142e8c505}{Appendix
3: Getting Started Checklist}}




\end{document}
